---
title: "working_final proj"
author: "Kaya Williams"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
library(here)
library(ggpmisc)
library(ggplot2)


nc<- read_csv(here("Final Project/OverallCounts_tidy.csv"))


nc<- nc[31:52,]

nc <- nc %>% 
  rename(`Covid_like_illness` = "COVID-like Illness",
         `RSV_like_illness` = "RSV-like Illness") %>% 
  separate(Week_of, c("Date","Year"), sep = ',') %>% 
  mutate(total = Covid_like_illness + RSV_like_illness + Influenza) %>% 
  select(-2)

nc$Date <- as.character(nc$Date)
nc$Date <- factor(nc$Date, levels=unique(nc$Date))

```


```{r, hurricane helene graph, total}


helene_illness <- ggplot(data = nc, aes(x = Date, y = total))+
  geom_line(linewidth = 1, na.rm = T, aes(group=1))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  labs(x = "Week",
       y = "Total Emergency Department Visits for Respitory Illnesses",
       title = "North Carolina Respitory Illness Trends in 2024")

helene_illness

helene_illness + annotate("rect", xmin = "September 21", xmax = "October 5", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "red") +
  annotate("rect", xmin = "October 5", xmax = "October 19", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "blue")+
  annotate("rect", xmin = "October 19", xmax = "November 2", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "green")
  
```
```{r, helene by illness}

nc_pivot <- nc %>% 
  pivot_longer(2:4,
               names_to = "Illness",
               values_to = "Count") %>% 
  select(-total)



(helene_by_illness <- ggplot(data = nc_pivot, aes(x = Date, y = Count, color = Illness))+
  geom_point(size = 1, na.rm = T)+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  labs(x = "Week",
       y = "Total Emergency Department Visits for Respitory Illnesses",
       title = "North Carolina Respitory Illness Trends in 2024")) 

helene_by_illness

helene_by_illness + annotate("rect", xmin = "September 21", xmax = "September 28", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "red") +
  annotate("rect", xmin = "September 28", xmax = "October 12", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "blue")+
  annotate("rect", xmin = "October 12", xmax = "October 26", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "green")

```



# Louisiana Hurricane Francine

```{r}
LA <- read_csv(here("Final Project/CSV_LA_COVID_TESTBYWEEK_TRACT_PUBLICUSE.csv"))

LA$Date_start_week <- as.character(LA$Date_start_week)
LA$Date_start_week <- factor(LA$Date_start_week, levels=unique(LA$Date_start_week))


LA <- LA %>% 
  filter(str_ends(Date_start_week, "/20")) #%>%
  #filter(!str_starts(Date_start_week, "1")) %>% 
  # filter(!str_starts(Date_start_week, "2")) %>% 
  # filter(!str_starts(Date_start_week, "3")) %>% 
  # filter(!str_starts(Date_start_week, "4")) %>% 
  # filter(!str_starts(Date_start_week, "5")) %>% 
  # select(-2,-3,-6,-7,-8) 


LA<- LA %>% 
  group_by(Date_start_week) %>% 
  summarise(Count_total = sum(Weekly_Case_Count))



LA_graph <- ggplot(data = LA, aes(x = Date_start_week, y = Count_total))+
  geom_line(linewidth = 1, na.rm = T, aes(group=1))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  labs(x = "Week",
       y = "Total Emergency Department Visits for Respitory Illnesses",
       title = "Louisiana Respitory Illness Trends in 2020")

LA_graph 
LA_graph_annotated <- LA_graph + annotate("rect", xmin = "8/27/20", xmax = "9/3/20", ymin = 0, ymax = 25000,
             alpha = 0.3, fill = "red") +
  annotate("rect", xmin = "9/3/20", xmax = "9/10/20", ymin = 0, ymax = 25000,
             alpha = 0.3, fill = "green")


LA_graph_annotated
```


Hurricane Helene - caused flooding as well 
Wind and flooding from Hurricane Helene devastated North Carolina on September 27, 2024, 
24-29

Hurricane laura - lousiana, august 27-29 2020



flood - july 2025 central texas flood, Guadalupe river 
West Virginia, June 2016 - 23 deaths



From January 7 to 31, 2025, a series of 14 destructive wildfires affected the Los Angeles metropolitan area and San Diego County in California, United States
Maui Wildfires, Hawaii (2023) August 2023




https://www.nbcdfw.com/news/national-international/deadliest-floods-in-the-us-in-the-last-25-years/3878935/

https://ncdetect.org/respiratory-dashboard/ 
https://covid19.ncdhhs.gov/dashboard

https://texas-respiratory-illness-dashboard-txdshsea.hub.arcgis.com/pages/texas-statewide-emergency-department-visits-for-respiratory-illnesses
https://texas-respiratory-illness-dashboard-txdshsea.hub.arcgis.com/pages/texas-statewide-emergency-department-visits-for-respiratory-illnesses






