---
title: "Impacts of Major Natural Disasters on American Communities"
author: "Alyssa Girard, Akillesh Suresh, Kaya Williams"
date: "2025-12-11"
output:
  html_document:
    toc: yes
    theme: cerulean
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
library(here)
library(dplyr)
library(treemapify)
library(ggpmisc)
library(ggplot2)
library(tigris)
library(viridis)
library(scales)
library(grid)
library(lubridate)
library(sf)
```

## Introduction
### How Do Hurricanes, Wildfires, and Floods Affect Human Welfare and Economics in the United States?
  Communities across the United States face increasing and ongoing disaster risks, resulting in impacts on how individuals live and the economic growth potential of those communities. This study aims to answer the question: How do hurricanes, wildfires, and floods affect human welfare and economics in the United States? These hazards represent the largest volume and most expensive disasters in our nation. As climate change increases their frequency and intensity, it is essential to assess the detrimental impacts they create. The information gained from examining the residual impacts of these natural disasters will help identify the areas in our country likely to be most adversely affected and the demographic groups that will bear the greatest risks. The results will provide the information needed to establish policies and plans to prepare for future natural disasters.

  To answer this complex question, we developed several follow-up questions that examine different dimensions of human and economic impact. We asked how rates of respiratory symptoms change after natural disasters, as disasters can disrupt air quality, healthcare access, and environmental conditions. Next, we evaluated how much natural disasters affect the U.S. economy by examining inflation-adjusted damages over time. Because damages vary widely across the country, we also asked which U.S. regions experience the highest economic losses, allowing us to assess geographic exposure and vulnerability. Finally, we investigated which disasters cause the most deaths and which result in the highest rates of homelessness, two key indicators of how disasters directly affect human lives and long-term community stability.

  Together, these follow-up questions allow us to examine the problem from multiple angles: health, mortality, homelessness, geography, and financial loss. By combining these perspectives, we can better understand how hurricanes, wildfires, and floods shape human welfare and economic resilience across the United States.

## Data & Data Sources
### Natural Disasters: Our World in Data
- Our first data source is from Our World in Data's Natural Disasters dataset
- It contains data from multiple types of disasters and impacts (e.g., deaths, injuries, homeless, et cetera) and the year the disaster occurred.
- We believe this data source is valid, because Our World In Data gets their data from valid sources, listed on their website as "international institutions and statistical agencies, such as the OECD, the World Bank, the WHO, and UN institutions, specialized institutes and think tanks, such as the Peace Research Institute Oslo (PRIO), academic research articles, such as Bourguignon & Morrison (2002) ’Inequality Among World Citizens: 1820-1992’ in the American Economic Review, and official data from government sources."
- Link: https://ourworldindata.org/natural-disasters#explore-data-on-natural-disasters

### Losses from Natural Disasters: Federal Reserve Bank of New York
- Our second data source is the Federal Reserve Bank of New York's Losses from Natural Disasters, which describes the direct damages from natural disasters across the United States by county.
- We believe this data source is valid because it is from a federal resource, specifically the Federal Reserve Bank of New York.
- Link: https://www.newyorkfed.org/research/policy/natural-disaster-losses/#interactive

### A Weekly Influenza Surveillance Report: CDC
- Our third data source is the CDC's Weekly Influenze Surveillance Report, which tracks the level of visits for respiratory illness.
- We believe this source is valid because it comes directly from the Center of Disease Control, a government organization.
- Link: https://gis.cdc.gov/grasp/fluview/main.html


## Results
### How do rates of respiratory symptoms change after natural disasters?
```{r}

```


### How much do natural disasters affect the U.S. economy?
```{r}

losses <- read_csv(here("data/lnd_full_clean_2024-12-31.csv"))

losses <- losses %>%
  mutate(
    begin_date = as.Date(begin_date),
    year = year(begin_date)
  )

selected_losses <- losses %>%
  filter(event_type %in% c("Hurricane", "Flood", "Tornado", "Wildfire"))

yearly_losses <- selected_losses %>%
  group_by(year, event_type) %>%
  summarise(total_damage = sum(damages_total_adj, na.rm = TRUE)) %>%
  ungroup()
```

```{r}
ggplot(yearly_losses, aes(
  x = year,
  y = total_damage,
  color = event_type
)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 1.8) +
  scale_y_continuous(
    labels = scales::label_number(scale_cut = scales::cut_short_scale())
  ) +
  facet_wrap(~ event_type, scales = "free_y") +
  labs(
  title = "Hurricanes Dominate U.S. Disaster-Related Economic Losses",
  subtitle = "Inflation-adjusted annual damages (1995–2024) reveal distinct trends across flood, 
  hurricane, tornado, and wildfire events",
  x = "Year",
  y = "Total Annual Damage (USD)"
) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

```


### Which U.S. regions experience the highest economic losses from natural disasters?
```{r}
options(tigris_use_cache = TRUE)

counties <- counties(cb = TRUE, year = 2021)

losses <- losses %>%
  mutate(fips = str_pad(fips, width = 5, side = "left", pad = "0")) %>%
  filter(damages_total_adj > 0)

major_types <- c(
  "Flood", "Flash Flood", "Hurricane", "Tropical Storm",
  "Storm Surge/Tide", "Tornado", "Wildfire", "High Wind",
  "Winter Storm", "Thunderstorm Wind"
)

losses_major <- losses %>%
  filter(event_type %in% major_types)

county_damage <- losses_major %>%
  group_by(fips) %>%
  summarize(total_damage = sum(damages_total_adj, na.rm = TRUE))

states <- states(cb = TRUE, year = 2021)

map_data <- counties %>%
  left_join(county_damage, by = c("GEOID" = "fips"))

map_data$total_damage[is.na(map_data$total_damage)] <- 0

```

```{r, fig.width=14, fig.height=10}
ggplot(map_data) +
  geom_sf(aes(fill = total_damage), color = NA) +
  geom_sf(data = states, fill = NA, color = "white", size = 0.3) +
  
  scale_fill_viridis_c(
    option = "A",
    trans = "log10",
    labels = scales::label_number(
      accuracy = 0.1,
      scale_cut = scales::cut_short_scale()
    )
  ) +
  labs(
    title = "Total Natural Disaster Damage by County",
    subtitle = "Cumulative inflation-adjusted damage from major hazard events",
    fill = "Total Damage"
  ) +
  coord_sf(
    xlim = c(-125, -65),
    ylim = c(25, 50),
    expand = FALSE
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(4, "cm"),
    legend.key.height = unit(0.7, "cm"),
    panel.grid = element_blank()
  )

```


### What disasters cause the most deaths?
```{r}
deaths <- read_csv(here("data/number-of-deaths-by-disaster.csv"))

deaths_usa <- deaths %>%
  filter(Entity == "United States") %>%
  select(Year,
         `Deaths - Flood`,
         `Deaths - Earthquake`,
         `Deaths - Extreme weather`)

long_data <- deaths_usa %>%
  pivot_longer(cols = -Year, names_to = "Disaster", values_to = "Deaths") %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0, Deaths))

total_data <- long_data %>%
  group_by(Disaster) %>%
  summarise(Total_Deaths = sum(Deaths))

total_data <- total_data %>%
  mutate(Disaster = recode(Disaster,
                                      "Deaths - Earthquake" = "Earthquake",
                                      	
"Deaths - Extreme weather" = "Extreme Weather",
"Deaths - Flood" = "Flood"
))

ggplot(data=total_data, aes(
  area = Total_Deaths,
  fill = Disaster,
  subgroup = Disaster,
  label = Total_Deaths
)) +
  geom_treemap() +
  geom_treemap_subgroup_border()+
  geom_treemap_subgroup_text(place = "centre",
                             size = 24,
                             alpha = 0.7,
                             colour="black",
                             fontface="italic")+
  geom_treemap_text(
    place = "topleft",
    colour = "black",
    grow = TRUE
  ) +
  scale_fill_manual(values = c(
    "Flood" = "lightblue",
    "Earthquake" = "pink",
    "Extreme Weather" = "orange"
  )) +
  labs(title = "Total Deaths by Disaster Type in the US",
       subtitle = "Floods, Earthquakes, and Extreme Weather (Tornadoes, Hurricanes, etc)")
```


### What disasters cause the highest numbers of homelessness?
```{r}
homeless <- read_csv(here("data/number-homeless-from-natural-disasters.csv"))

three_types <- homeless %>%
  group_by(Entity) %>%
  summarise(total_homeless = sum(Homeless, na.rm = TRUE)) %>%
  filter(Entity %in% c("Extreme weather", "Flood", "Wildfire")) %>%
  mutate(Disaster = Entity)

all_disasters <- homeless %>%
  group_by(Entity) %>%
  summarise(total_homeless = sum(Homeless, na.rm = TRUE)) %>%
  filter(Entity == "All disasters") %>%
  mutate(Disaster = Entity)

ggplot(data = three_types,
       aes(x = Disaster,
           y = total_homeless / 1000,
           fill = Disaster))+
  geom_bar(stat = "identity") +
  geom_text(aes(label = total_homeless/10000, vjust = -0.5)) +
  scale_y_continuous(limits = c(0, 125000)) +
  labs(title = "Cases of Homelessness by Disaster (in tens of thousands)",
       x = "Type of Disaster",
       y = "Cases (tens of thousands)")
```


## Conclusion

## Contributions
### Kaya Williams
 - How do rates of respiratory symptoms change after natural disasters?
 charts

### Akillesh Suresh
 - Wrote the introduction paragraphs
 - Which U.S. regions experience the highest economic losses from natural disasters? chart
 - How much do natural disasters affect the U.S. economy? chart

### Alyssa Girard
 - What disasters cause the most deaths? chart
 - What disasters cause the highest rates of homelessness? chart