---
title: "How Do Hurricanes, Wildfires, and Floods Affect Human Welfare and Economics in the United States?"
author: "Kaya Williams, Akillesh Suresh, Alyssa Girard"
date: "2025-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(here)
library(dplyr)
library(treemapify)
library(ggpmisc)
library(ggplot2)
library(tigris)
library(viridis)
library(scales)
library(grid)
library(lubridate)
library(sf)
```

## Data Sources
### Natural Disasters: Our World in Data
- Our first data source is from Our World in Data's Natural Disasters dataset
- It contains data from multiple types of disasters and impacts (e.g., deaths, injuries, homeless, et cetera) and the year the disaster occurred.
- We believe this data source is valid, because Our World In Data gets their data from valid sources, listed on their website as "international institutions and statistical agencies, such as the OECD, the World Bank, the WHO, and UN institutions, specialized institutes and think tanks, such as the Peace Research Institute Oslo (PRIO), academic research articles, such as Bourguignon & Morrison (2002) ’Inequality Among World Citizens: 1820-1992’ in the American Economic Review, and official data from government sources."
- Link: https://ourworldindata.org/natural-disasters#explore-data-on-natural-disasters

### Losses from Natural Disasters: Federal Reserve Bank of New York
- Our second data source is the Federal Reserve Bank of New York's Losses from Natural Disasters, which describes the direct damages from natural disasters across the United States by county.
- We believe this data source is valid because it is from a federal resource, specifically the Federal Reserve Bank of New York.
- Link: https://www.newyorkfed.org/research/policy/natural-disaster-losses/#interactive

### A Weekly Influenza Surveillance Report: CDC
- Our third data source is the CDC's Weekly Influenze Surveillance Report, which tracks the level of visits for respiratory illness.
- We believe this source is valid because it comes directly from the Center of Disease Control, a government organization.
- Link: https://gis.cdc.gov/grasp/fluview/main.html


## Evaluating Our Proposal Expectations

### How do rates of respiratory symptoms change after natural disasters?

### How much do natural disasters affect the U.S. economy?

### Which U.S. regions experience the highest economic losses from natural disasters?

### What disasters cause the most deaths?
We originally asked the question "What disasters cause people to require the most assistance", but based on the feedback we received it was far too vague. In order to be more specific in our research, we decided to change the question entirely to "what disasters cause the most deaths?" For our original question, we predicted that floods would cause the most need for assistance. Translating this for our new question into predicting that floods cause the most deaths, we saw that extreme weather (i.e. hurricanes and tornadoes) caused the most, with 31,825 deaths while floods had 3,463. 

###  What disasters cause the highest rates of homelessness?
Our question was "What disasters cause the highest rates of homelessness?" and we predicted that the disasters that would cause the highest rates of homelessness are wildfires and floods. In our research we saw that floods indeed caused the most, with around 9478 tens of thousands of cases of homelessness, but extreme weather, with 5643 tens of thousands of cases, followed with wildfires being in last with 25 tens of thousands of cases. 


## Preliminary Charts

### How do rates of respiratory symptoms change after natural disasters?
```{r}
nc<- read_csv(here("data/OverallCounts_tidy.csv"))


nc<- nc[31:52,]

nc <- nc %>% 
  rename(`Covid_like_illness` = "COVID-like Illness",
         `RSV_like_illness` = "RSV-like Illness") %>% 
  separate(Week_of, c("Date","Year"), sep = ',') %>% 
  mutate(total = Covid_like_illness + RSV_like_illness + Influenza) %>% 
  select(-2)

nc$Date <- as.character(nc$Date)
nc$Date <- factor(nc$Date, levels=unique(nc$Date))

```

```{r, hurricane helene graph, total}


helene_illness <- ggplot(data = nc, aes(x = Date, y = total))+
  geom_line(linewidth = 1, na.rm = T, aes(group=1))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  labs(x = "Week",
       y = "Total Emergency Department Visits for Respitory Illnesses",
       title = "North Carolina Respitory Illness Trends in 2024")

helene_illness

p1<- helene_illness + annotate("rect", xmin = "September 21", xmax = "October 5", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "red") +
  annotate("rect", xmin = "October 5", xmax = "October 19", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "blue")+
  annotate("rect", xmin = "October 19", xmax = "November 2", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "green")

p1  
```



```{r, helene by illness}

nc_pivot <- nc %>% 
  pivot_longer(2:4,
               names_to = "Illness",
               values_to = "Count") %>% 
  select(-total)



(helene_by_illness <- ggplot(data = nc_pivot, aes(x = Date, y = Count, color = Illness))+
  geom_point(size = 1, na.rm = T)+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  labs(x = "Week",
       y = "Total Emergency Department Visits for Respitory Illnesses",
       title = "North Carolina Respitory Illness Trends in 2024")) 

helene_by_illness

helene_by_illness + annotate("rect", xmin = "September 21", xmax = "September 28", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "red") +
  annotate("rect", xmin = "September 28", xmax = "October 12", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "blue")+
  annotate("rect", xmin = "October 12", xmax = "October 26", ymin = 0, ymax = 12000,
             alpha = 0.3, fill = "green")

```
```{r}
LA <- read_csv(here("data/CSV_LA_COVID_TESTBYWEEK_TRACT_PUBLICUSE.csv"))

LA$Date_start_week <- as.character(LA$Date_start_week)
LA$Date_start_week <- factor(LA$Date_start_week, levels=unique(LA$Date_start_week))


LA <- LA %>% 
  filter(str_ends(Date_start_week, "/20")) #%>%
  #filter(!str_starts(Date_start_week, "1")) %>% 
  # filter(!str_starts(Date_start_week, "2")) %>% 
  # filter(!str_starts(Date_start_week, "3")) %>% 
  # filter(!str_starts(Date_start_week, "4")) %>% 
  # filter(!str_starts(Date_start_week, "5")) %>% 
  # select(-2,-3,-6,-7,-8) 


LA<- LA %>% 
  group_by(Date_start_week) %>% 
  summarise(Count_total = sum(Weekly_Case_Count))



LA_graph <- ggplot(data = LA, aes(x = Date_start_week, y = Count_total))+
  geom_line(linewidth = 1, na.rm = T, aes(group=1))+
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  labs(x = "Week",
       y = "Total Emergency Department Visits for Respitory Illnesses",
       title = "Louisiana Respitory Illness Trends in 2020")

LA_graph 
LA_graph_annotated <- LA_graph + annotate("rect", xmin = "8/27/20", xmax = "9/3/20", ymin = 0, ymax = 25000,
             alpha = 0.3, fill = "red") +
  annotate("rect", xmin = "9/3/20", xmax = "9/10/20", ymin = 0, ymax = 25000,
             alpha = 0.3, fill = "green")


LA_graph_annotated
```

### How much do natural disasters affect the U.S. economy?

```{r}

losses <- read_csv(here("data/lnd_full_clean_2024-12-31.csv"))

losses <- losses %>%
  mutate(
    begin_date = as.Date(begin_date),
    year = year(begin_date)
  )

selected_losses <- losses %>%
  filter(event_type %in% c("Hurricane", "Flood", "Tornado", "Wildfire"))
```

```{r}
ggplot(selected_losses, aes(x = year, 
                            y = damages_total_adj,
                            color = event_type, 
                            group = event_type)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  labs(
    title = "Economic Losses Over Time by Natural Disaster Type",
    x = "Year",
    y = "Inflation-Adjusted Damage (USD)",
    color = "Disaster Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 14)
  )
```

### Which U.S. regions experience the highest economic losses from natural disasters?

```{r}
options(tigris_use_cache = TRUE)

counties <- counties(cb = TRUE, year = 2021)

losses <- read_csv(here("data/lnd_full_clean_2024-12-31.csv"))

losses <- losses %>%
  mutate(fips = str_pad(fips, width = 5, side = "left", pad = "0")) %>%
  filter(damages_total_adj > 0)

major_types <- c(
  "Flood", "Flash Flood", "Hurricane", "Tropical Storm",
  "Storm Surge/Tide", "Tornado", "Wildfire", "High Wind",
  "Winter Storm", "Thunderstorm Wind"
)

losses_major <- losses %>%
  filter(event_type %in% major_types)

county_damage <- losses_major %>%
  group_by(fips) %>%
  summarize(total_damage = sum(damages_total_adj, na.rm = TRUE))

map_data <- counties %>%
  left_join(county_damage, by = c("GEOID" = "fips"))

map_data$total_damage[is.na(map_data$total_damage)] <- 0

```

```{r, fig.width=14, fig.height=10}
ggplot(map_data) +
  geom_sf(aes(fill = total_damage), color = NA) +
  scale_fill_viridis_c(
    option = "A",
    trans = "log10",
    labels = scales::label_number(
      accuracy = 0.1,
      scale_cut = scales::cut_short_scale()
    )
  ) +
  labs(
    title = "Total Natural Disaster Damage by County",
    subtitle = "Cumulative inflation-adjusted damage from major hazard events",
    fill = "Total Damage"
  ) +
  coord_sf(
    xlim = c(-125, -65),
    ylim = c(25, 50),
    expand = FALSE
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(4, "cm"),
    legend.key.height = unit(0.7, "cm"),
    panel.grid = element_blank()
  )
```

### What disasters cause the most deaths?
```{r}
# import data
deaths <- read_csv(here("data/number-of-deaths-by-disaster.csv"))

deaths_usa <- deaths %>%
  filter(Entity == "United States") %>%
  select(Year,
         `Deaths - Flood`,
         `Deaths - Earthquake`,
         `Deaths - Extreme weather`)

long_data <- deaths_usa %>%
  pivot_longer(cols = -Year, names_to = "Disaster", values_to = "Deaths") %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0, Deaths))

total_data <- long_data %>%
  group_by(Disaster) %>%
  summarise(Total_Deaths = sum(Deaths))

ggplot(total_data, aes(
  area = Total_Deaths,
  fill = Disaster,
  label = paste0(Disaster, "\n", Total_Deaths)
)) +
  geom_treemap() +
  geom_treemap_text(
    colour = "white",
    place = "centre",
    grow = TRUE
  ) +
  scale_fill_manual(values = c(
    "Deaths - Flood" = "blue",
    "Deaths - Earthquake" = "red",
    "Deaths - Extreme weather" = "orange"
  )) +
  labs(title = "Total Deaths by Disaster Type") +
  theme(legend.position = "none")
```

### What disasters cause the highest rates of homelessness?
```{r}
homeless <- read_csv(here("data/number-homeless-from-natural-disasters.csv"))

three_types <- homeless %>%
  group_by(Entity) %>%
  summarise(total_homeless = sum(Homeless, na.rm = TRUE)) %>%
  filter(Entity %in% c("Extreme weather", "Flood", "Wildfire"))

all_disasters <- homeless %>%
  group_by(Entity) %>%
  summarise(total_homeless = sum(Homeless, na.rm = TRUE)) %>%
  filter(Entity == "All disasters")

ggplot(data = three_types,
       aes(x = Entity, y = total_homeless / 1000))+
  geom_bar(stat = "identity", fill = "blue") +
  geom_text(aes(label = total_homeless/10000, vjust = -0.5, size = 1)) +
  scale_y_continuous(limits = c(0, 125000)) +
  labs(title = "",
       x = "Type of Disaster",
       y = " (in tens of thousands)")
```


## Member Contributions
### Kaya Williams
 - How do rates of respiratory symptoms change after natural disasters?
 charts
 - Worked on progress report file
 
### Akillesh Suresh
 - Which U.S. regions experience the highest economic losses from natural disasters? chart
 - How much do natural disasters affect the U.S. economy? chart
 - Worked on progress report file

### Alyssa Girard
 - What disasters cause the most deaths? chart
 - What disasters cause the highest rates of homelessness? chart
 - Worked on progress report file
